---
title: "AMAZE- Pathway Analysis"
Author: Monica Mweetwa
Date: 20th May 2024
---

1. Set working directory
```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("C:/Users/Monica/Dropbox/AMAZE")) 
    #knitr::opts_knit$set(root.dir = normalizePath("~/Library/CloudStorage/Dropbox/AMAZE")) 
```

2. Load necessary packages 
```{r include=FALSE}
#BiocManager::install("MicrobiomeProfiler")
library(MicrobiomeProfiler)
library(ggpicrust2)
library(readr)
library(Maaslin2)
```

This analysis take output from picrust2 as input that was run in a conda environment (picrust-src) using the follwoing command:
picrust2_pipeline.py -s ~/Dropbox/AMAZE/Inputs/Picrust2/AMAZE_sequences.fna \
-i ~/Dropbox/AMAZE/Inputs/Picrust2/AMAZE_biom.biom \
-o ~/Dropbox/AMAZE/Outputs/G_Picrust2/AMAZE_out_picrust -p 1

This produced metagenome and pathway predictions based on the abundance of 16 taxa in our dataset. This can be analysed similar to taxonomic abundance analysis using ggpicrust2

Implement:

Import data
```{r}
#If you want to analysis the EC. MetaCyc. KO without conversions. You should turn ko_to_kegg to FALSE.
#metadata
metadata <- read_csv("Inputs/Metadata/metadata_paired_withELISA.csv") %>%
  column_to_rownames(var = "SampleID2")
levels(as.factor(metadata$InterventionArm))
metadata$Intervention_arm <- relevel(as.factor(metadata$InterventionArm), "Placebo")

#pathways
pathway_dat <- read.delim("Outputs/4. Picrust2/AMAZE_out_picrust/pathways_out/path_abun_unstrat.tsv")%>%
  column_to_rownames(var = "pathway")
names(pathway_dat) <- substring(names(pathway_dat),2)

#Get pathway descriptions and create feature.ann
library(ggpicrust2)
path_df <- pathway_annotation(
  file = "Outputs/4. Picrust2/AMAZE_out_picrust/pathways_out/path_abun_unstrat.tsv",
  pathway = "MetaCyc",
  daa_results_df = path_df,
  ko_to_kegg = FALSE
)

#add pathway description to file
library(tidyverse)
feature_ann <- path_df %>% select(pathway, description) %>% mutate(pathID = pathway) %>%column_to_rownames(var = "pathway")
```

Create models
```{r}
mas_P <- Maaslin2(
  input_data = pathway_dat,
  input_metadata = metadata %>% filter(Sample.Type == "Duodenal Aspirate"),
  output = "Outputs/3. SupplementationAnalysis/PathwayAnalysis/maaslin2",
  min_abundance = 0.001,
  min_prevalence = 0.05,
  normalization = "TMM",
  transform = "NONE",
  analysis_method = "NEGBIN",
  max_significance = 0.1,
  fixed_effects = c("InterventionArm", "Timepoint"),
  random_effects = "PID",
  correction = "BY",
  standardize = FALSE,
  cores = 1,
  plot_heatmap = TRUE,
  plot_scatter = TRUE,
  heatmap_first_n = 50,
  reference = ('InterventionArm,Placebo'))
res <- mas_P[["results"]]
notsig <- res %>%
  mutate(sam.type = ifelse(.$coef > 0,  "Stool", "Duodenal Aspirate"),
         sam.type = (ifelse(.$qval > 0.05 , "Common", sam.type)),
         coef.exp = exp(coef),
         logfold2 = log2(coef.exp),
         logfold2.2 = ifelse(logfold2 < 0, logfold2*-1, logfold2)) 
#replace the dots with high-phen
notsig2 <- notsig %>%
  mutate(feature = str_replace_all(feature, "\\.", "-"))

write_csv(notsig2, "Outputs/3. SupplementationAnalysis/PathwayAnalysis/lda_pathway_intervention.csv")
```



Placebo as reference
```{r}
library(MicrobiomeStat)

#1. duodenal samples
metadata_duod <- metadata %>% filter(Sample.Type == "Duodenal Aspirate")
MicrobiomeData_duod <- list(
  feature.tab = as.matrix(pathway_dat),
  meta.dat = metadata_duod,
  feature.ann = as.matrix(feature_ann)
)

#add time variable with pre-intervention as 0
MicrobiomeData_duod[["meta.dat"]]$InterventionArm3 <- ifelse(MicrobiomeData_duod[["meta.dat"]]$InterventionArm == "Placebo", "_Placebo", MicrobiomeData_duod[["meta.dat"]]$InterventionArm)
#Create models : pathway change test pair
test.list_duod <- generate_taxa_change_test_pair(
  data.obj = MicrobiomeData_duod,
  subject.var = "PID",
  time.var = "Timepoint",
  group.var = "InterventionArm3",
  adj.vars = NULL,
  change.base = "Pre-Intervention",
  feature.change.func = "log fold change",
  feature.level = "pathID",
  prev.filter = 0.01,
  abund.filter = 0.001,
  feature.dat.type = "count"
)

#No significant pathways between intervention arms

#2. stool samples
metadata_stool <- metadata %>% filter(Sample.Type == "Stool")
MicrobiomeData_stool <- list(
  feature.tab = as.matrix(pathway_dat),
  meta.dat = metadata_stool,
  feature.ann = as.matrix(feature_ann)
)
#add time variable with pre-intervention as 0
MicrobiomeData_stool[["meta.dat"]]$InterventionArm3 <- ifelse(MicrobiomeData_stool[["meta.dat"]]$InterventionArm == "Placebo", "_Placebo", MicrobiomeData_stool[["meta.dat"]]$InterventionArm)

#Create models : pathway change test pair
test.list_stool <- generate_taxa_change_test_pair(
  data.obj = MicrobiomeData_stool,
  subject.var = "PID",
  time.var = "Timepoint",
  group.var = "InterventionArm3",
  adj.vars = NULL,
  change.base = "Pre-Intervention",
  feature.change.func = "log fold change",
  feature.level = "pathID",
  prev.filter = 0.01,
  abund.filter = 0.001,
  feature.dat.type = "count"
)

#Miconutrient had 1 pathway significant after intervention compared to Placebo
#save results:
#save(test.list_duod, test.list_stool, file = "Outputs/3. SupplementationAnalysis/LDA_Pathway.RData")

#library(writexl)
#write.xlsx(test.list_duod, "Outputs/3. SupplementationAnalysis/LDA_Pathway_duod.xlsx")
#write.xlsx(test.list_stool, "Outputs/3. SupplementationAnalysis/LDA_Pathway_stool.xlsx")


```
Mironutrient as reference
```{r}
library(MicrobiomeStat)

#1. duodenal samples
metadata_duod <- metadata %>% filter(Sample.Type == "Duodenal Aspirate")
MicrobiomeData_duod <- list(
  feature.tab = as.matrix(pathway_dat),
  meta.dat = metadata_duod,
  feature.ann = as.matrix(feature_ann)
)

#add time variable with pre-intervention as 0
MicrobiomeData_duod[["meta.dat"]]$InterventionArm3 <- ifelse(MicrobiomeData_duod[["meta.dat"]]$InterventionArm == "Micronutrient", "_Micronutrient", MicrobiomeData_duod[["meta.dat"]]$InterventionArm)
#Create models : pathway change test pair
test.list_duod_mm <- generate_taxa_change_test_pair(
  data.obj = MicrobiomeData_duod,
  subject.var = "PID",
  time.var = "Timepoint",
  group.var = "InterventionArm3",
  adj.vars = NULL,
  change.base = "Pre-Intervention",
  feature.change.func = "log fold change",
  feature.level = "pathID",
  prev.filter = 0.01,
  abund.filter = 0.001,
  feature.dat.type = "count"
)

#No significant pathways between intervention arms

#2. stool samples
metadata_stool <- metadata %>% filter(Sample.Type == "Stool")
MicrobiomeData_stool <- list(
  feature.tab = as.matrix(pathway_dat),
  meta.dat = metadata_stool,
  feature.ann = as.matrix(feature_ann)
)
#add time variable with pre-intervention as 0
MicrobiomeData_stool[["meta.dat"]]$InterventionArm3 <- ifelse(MicrobiomeData_stool[["meta.dat"]]$InterventionArm == "Micronutrient", "_Micronutrient", MicrobiomeData_stool[["meta.dat"]]$InterventionArm)

#Create models : pathway change test pair
test.list_stool_mm <- generate_taxa_change_test_pair(
  data.obj = MicrobiomeData_stool,
  subject.var = "PID",
  time.var = "Timepoint",
  group.var = "InterventionArm3",
  adj.vars = NULL,
  change.base = "Pre-Intervention",
  feature.change.func = "log fold change",
  feature.level = "pathID",
  prev.filter = 0.01,
  abund.filter = 0.001,
  feature.dat.type = "count"
)
```

Combined as reference
```{r}
library(MicrobiomeStat)

#1. duodenal samples
metadata_duod <- metadata %>% filter(Sample.Type == "Duodenal Aspirate")
MicrobiomeData_duod <- list(
  feature.tab = as.matrix(pathway_dat),
  meta.dat = metadata_duod,
  feature.ann = as.matrix(feature_ann)
)

#add time variable with pre-intervention as 0
MicrobiomeData_duod[["meta.dat"]]$InterventionArm3 <- ifelse(MicrobiomeData_duod[["meta.dat"]]$InterventionArm == "Combined", "_Micronutrient", MicrobiomeData_duod[["meta.dat"]]$InterventionArm)
#Create models : pathway change test pair
test.list_duod_cc <- generate_taxa_change_test_pair(
  data.obj = MicrobiomeData_duod,
  subject.var = "PID",
  time.var = "Timepoint",
  group.var = "InterventionArm3",
  adj.vars = NULL,
  change.base = "Pre-Intervention",
  feature.change.func = "log fold change",
  feature.level = "pathID",
  prev.filter = 0.01,
  abund.filter = 0.001,
  feature.dat.type = "count"
)

#No significant pathways between intervention arms

#2. stool samples
metadata_stool <- metadata %>% filter(Sample.Type == "Stool")
MicrobiomeData_stool <- list(
  feature.tab = as.matrix(pathway_dat),
  meta.dat = metadata_stool,
  feature.ann = as.matrix(feature_ann)
)
#add time variable with pre-intervention as 0
MicrobiomeData_stool[["meta.dat"]]$InterventionArm3 <- ifelse(MicrobiomeData_stool[["meta.dat"]]$InterventionArm == "Combined", "_Combined", MicrobiomeData_stool[["meta.dat"]]$InterventionArm)

#Create models : pathway change test pair
test.list_stool_cc <- generate_taxa_change_test_pair(
  data.obj = MicrobiomeData_stool,
  subject.var = "PID",
  time.var = "Timepoint",
  group.var = "InterventionArm3",
  adj.vars = NULL,
  change.base = "Pre-Intervention",
  feature.change.func = "log fold change",
  feature.level = "pathID",
  prev.filter = 0.01,
  abund.filter = 0.001,
  feature.dat.type = "count"
)
```

Generate volcano plots
```{r}
# Stool
plot.list <- generate_taxa_volcano_single(
  data.obj = MicrobiomeData_stool,
  group.var = "InterventionArm3",
  test.list = test.list_stool,
  feature.sig.level = 0.05,
  feature.mt.method = "fdr"
)
#duodenum
plot.list1 <- generate_taxa_volcano_single(
  data.obj = MicrobiomeData_duod,
  group.var = "InterventionArm3",
  test.list = test.list_duod,
  feature.sig.level = 0.05,
  feature.mt.method = "fdr"
)
library(ggpubr)
g1 <- ggarrange(ggarrange(plot.list1[["pathID"]][["Amino Acid vs _Placebo (Reference)"]],
                            plot.list1[["pathID"]][["Combined vs _Placebo (Reference)"]],
                            plot.list1[["pathID"]][["Micronutrient vs _Placebo (Reference)"]],
                            nrow = 1,
                          common.legend = T,
                          legend = "right"),
                ggarrange(plot.list[["pathID"]][["Amino Acid vs _Placebo (Reference)"]],
                          plot.list[["pathID"]][["Combined vs _Placebo (Reference)"]],
                          plot.list[["pathID"]][["Micronutrient vs _Placebo (Reference)"]],
                          nrow = 1,
                          common.legend = T,
                          legend = "right"),
                nrow = 2,
                common.legend = T,
                legend = "right")

ggsave(plot = g1, filename = "Outputs/3. SupplementationAnalysis/PathwayAnalysis/ALLvsPlacebo2.png", dpi = 600, width = 45, height = 20, units = "cm")
```

Indivdual pathways boxplot
```{r}
generate_taxa_indiv_boxplot_long(
   data.obj = MicrobiomeData_stool,
   subject.var = "PID",
   time.var = "Timepoint",
   t0.level = "Pre-Intervention",
   ts.levels = "Post-Intervention",
   group.var = "InterventionArm",
   #strata.var = "sex",
   feature.level = c("pathID"),
   feature.dat.type = "count",
   features.plot = NULL,
   top.k.plot = NULL,
   top.k.func = NULL,
   transform = "log",
   prev.filter = 0.1,
   abund.filter = 0.001,
   base.size = 20,
   theme.choice = "bw",
   custom.theme = NULL,
   palette = NULL,
   pdf = TRUE,
   file.ann = NULL,
   pdf.wid = 11,
   pdf.hei = 8.5
)
```

