---
title: "AMAZE - Baseline Characteristics"
Author: Monica N Mweetwa
Date: 17/04/2024
---

1. Set working directory
```{r setup}
   # knitr::opts_knit$set(root.dir = normalizePath("C:/Users/Monica/Dropbox/AMAZE")) 
    knitr::opts_knit$set(root.dir = normalizePath("~/Library/CloudStorage/Dropbox/AMAZE")) 
```

2. Load necessary packages 
```{r include=FALSE}
#library(btools)
library(ggpubr) 
library(ggh4x)
library(ggplot2)
library(ggbeeswarm)
library(microViz)
library(phyloseq)
library(patchwork)
library(tidyverse)
```

3. Load data
```{r}
load("Inputs/RData/AMAZE_ps_filtered.RData")
ps_filtered
```

############################ baseline characteristics ################################
A. Relative abundance - phylum
```{r}
#1. Aspirates
ps_asp <- subset_samples(ps_filtered, Sample.Type == "Duodenal Aspirate")
ps_asp.pre <- subset_samples(ps_asp, Timepoint == "Pre-Intervention")
# Transform to relative abundance
AID_norm <- transform_sample_counts(ps_asp.pre, function(x) 100 * x/sum(x))
AID_Orders <- AID_norm  %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate taxa at order level
  psmelt() %>%                                        # Melt phyloseq object to long format for producing graphics with ggplot2
  filter(Abundance > 1.0)  %>%                        # Filter out phyla below 1% in each sample
  arrange(desc(Phylum))
# Sum remaining taxa with a relative abundance < 1% and make a new dataframe
Remainders <- (AID_Orders) %>%
  group_by(PID) %>% 
  summarise(Abundance = (100-sum(Abundance))) %>% 
  as.data.frame()
Remainders$Order<-"Phyla < 1%"
Remainders$Phylum<-"_Phyla < 1%"
Remainders$Class<-"_Phyla < 1%"
Remainders$Kingdom<-"Bacteria"
# Join dataframes
AID_barchart <- full_join(AID_Orders,Remainders)
# reorder based on phylogeny
AID_barchart_asp <- AID_barchart[with(AID_barchart, order(Phylum)),]

#2. Stool
ps_stool <- subset_samples(ps_filtered, Sample.Type == "Stool")
ps_stool.pre <- subset_samples(ps_stool, Timepoint == "Pre-Intervention")
# Transform to relative abundance
AID_norm <- transform_sample_counts(ps_stool.pre, function(x) 100 * x/sum(x))
AID_Orders <- AID_norm  %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate taxa at order level
  psmelt() %>%                                        # Melt phyloseq object to long format for producing graphics with ggplot2
  filter(Abundance > 1.0)  %>%                        # Filter out orders below 1% in each sample
  arrange(desc(Phylum))
# Sum remaining taxa with a relative abundance < 1% and make a new dataframe
Remainders <- (AID_Orders) %>%
  group_by(PID) %>% 
  summarise(Abundance = (100-sum(Abundance))) %>% 
  as.data.frame()
Remainders$Order<-"Phyla < 1%"
Remainders$Phylum<-"_Phyla < 1%"
Remainders$Class<-"_Phyla < 1%"
Remainders$Kingdom<-"Bacteria"
# Join dataframes
AID_barchart <- full_join(AID_Orders,Remainders)
# reorder based on phylogeny
AID_barchart_stool <- AID_barchart[with(AID_barchart, order(Phylum)),]

#3. plot
palette1 <- c("_Phyla < 1%" = "#808080", 
          "Actinobacteriota" = "#E41A1C", 
          "Bacteroidota" = "#FF8C00", 
          "Campilobacterota" = "#4DAF4A",  
          "Desulfobacterota" = "#984EA3", 
          "Firmicutes" = "#FFC0CB", 
          "Fusobacteriota" = "#40E0D0", 
          "Proteobacteria" = "#990000", 
          "Elusimicrobiota" = "#4B0082", 
          "Spirochaetota" = "#003380")

asp.plot <- ggplot(AID_barchart_asp, aes(x = factor(PID), y = Abundance, fill = Phylum)) +
  geom_bar(position="stack", stat="identity") +
  labs(x = "Samples", y = "Relative Abundance", title = "A) DUODENAL ASPIRATE") +
  scale_fill_manual(values = palette1) +
  theme_classic2() + 
  coord_flip() +
  theme(axis.ticks.y = element_blank())

stool.plot <- ggplot(AID_barchart_stool, aes(x = factor(PID), y = Abundance, fill = Phylum)) +
  geom_bar(position="stack", stat="identity") +
  labs(x = "Samples", y = "Relative Abundance",  title = "B) STOOL") +
  scale_fill_manual(values = palette1) +
  theme_classic2() + 
  coord_flip() +
  theme(axis.ticks.y = element_blank())

#combine and save
library(ggpubr)
baseline_relplot <- ggarrange(asp.plot, stool.plot, 
                              common.legend = TRUE, ncol = 1)

ggsave("Outputs/2. Baseline/SuppFig3.png", plot = baseline_relplot, dpi = 600, device = "png", width = 45, height = 35, units = "cm")
save(asp.plot, stool.plot, file = "Inputs/RData/ForCombinedPlots/Baseline_phyla.RData")
```

2. Relative abundance - Genus
```{r}
#1. Aspirates
ps_asp <- subset_samples(ps_filtered, Sample.Type == "Duodenal Aspirate")
ps_asp.pre <- subset_samples(ps_asp, Timepoint == "Pre-Intervention")
# Transform to relative abundance
AID_norm <- transform_sample_counts(ps_asp.pre, function(x) 100 * x/sum(x))
AID_Orders <- AID_norm  %>%
  tax_glom(taxrank = "Genus") %>%                     # agglomerate taxa at order level
  psmelt() %>%                                        # Melt phyloseq object to long format for producing graphics with ggplot2
  filter(Abundance > 20)  %>%                        # Filter out phyla below 1% in each sample
  arrange(desc(Genus))
# Sum remaining taxa with a relative abundance < 1% and make a new dataframe
Remainders <- (AID_Orders) %>%
  group_by(PID) %>% 
  summarise(Abundance = (100-sum(Abundance))) %>% 
  as.data.frame()
Remainders$Genus<-"-Genus < 20%"
Remainders$Family<-"Genus < 1%"
Remainders$Order<-"Genus < 1%"
Remainders$Phylum<-"_Genus < 1%"
Remainders$Class<-"_Genus < 1%"
Remainders$Kingdom<-"Bacteria"
# Join dataframes
AID_barchart <- full_join(AID_Orders,Remainders)
# reorder based on phylogeny
AID_barchart_asp <- AID_barchart[with(AID_barchart, order(Phylum)),]

#2. Stool
ps_stool <- subset_samples(ps_filtered, Sample.Type == "Stool")
ps_stool.pre <- subset_samples(ps_stool, Timepoint == "Pre-Intervention")
# Transform to relative abundance
AID_norm <- transform_sample_counts(ps_stool.pre, function(x) 100 * x/sum(x))
AID_Orders <- AID_norm  %>%
  tax_glom(taxrank = "Genus") %>%                     # agglomerate taxa at order level
  psmelt() %>%                                        # Melt phyloseq object to long format for producing graphics with ggplot2
  filter(Abundance > 20)  %>%                        # Filter out orders below 1% in each sample
  arrange(desc(Genus))
# Sum remaining taxa with a relative abundance < 1% and make a new dataframe
Remainders <- (AID_Orders) %>%
  group_by(PID) %>% 
  summarise(Abundance = (100-sum(Abundance))) %>% 
  as.data.frame()
Remainders$Genus<-"-Genus < 20%"
Remainders$Family<-"Genus < 1%"
Remainders$Order<-"Genus < 1%"
Remainders$Phylum<-"_Genus < 1%"
Remainders$Class<-"_Genus < 1%"
Remainders$Kingdom<-"Bacteria"
# Join dataframes
AID_barchart <- full_join(AID_Orders,Remainders)
# reorder based on phylogeny
AID_barchart_stool <- AID_barchart[with(AID_barchart, order(Phylum)),]

#3. plot
palette2 <- c('Alloprevotella' = "#E41A1C", 
              'Bacteroides' = "#FF8C00",
              'Burkholderia-Caballeronia-Paraburkholderia' ="#4DAF4A", 
              'Corynebacterium' = "#FFC0CB",
              'Escherichia-Shigella' = "#4B0082",
              'Faecalibacterium' = "#EE82EE", 
              'Gemella' = "#00FFFF", 
              '-Genus < 20%' ="gray", 
              'Haemophilus' ="#8F7C00", 
              'Methylobacterium-Methylorubrum' ="#003380",
              'Neisseria' = "#426600", 
              'Peptostreptococcus' ="#FF00FF", 
              'Prevotella' = "#FFFF00",
              'Ralstonia' ="#740AFF", 
              'Streptococcus' ="#990000",
              'Succinivibrio' = "#FF6A6A")
            
asp.plot <- ggplot(AID_barchart_asp, aes(x = factor(PID), y = Abundance, fill = Genus)) +
  geom_bar(position="stack", stat="identity") +
  labs(x = "Samples", y = "Relative Abundance", title = "A) DUODENAL ASPIRATE") +
  scale_fill_manual(values = palette2) +
  theme_classic2() + 
  coord_flip() +
  theme(axis.ticks.y = element_blank())

stool.plot <- ggplot(AID_barchart_stool, aes(x = factor(PID), y = Abundance, fill = Genus)) +
  geom_bar(position="stack", stat="identity") +
  labs(x = "Samples", y = "Relative Abundance",  title = "B) STOOL") +
  scale_fill_manual(values = palette2) +
  theme_classic2() + 
  coord_flip() +
  theme(axis.ticks.y = element_blank())

#combine and save
library(ggpubr)
baseline_relplot <- ggarrange(asp.plot, stool.plot, common.legend = TRUE, legend = "bottom",ncol = 2)

ggsave("Outputs/2. Baseline/Fig1.png", plot = baseline_relplot, dpi = 600, device = "png", width = 45, height = 35, units = "cm")
save(asp.plot, stool.plot, file = "Inputs/RData/ForCombinedPlots/Baseline_genus.RData")
```

3. Alpha Diversity - Faiths PD metric association with sample characteristics
```{r}
#function to create linear regression models
lm_uni_func = function(equation,data){
  lm_mod = lm(formula = equation, data = data) #family must be binomial if y-values are between 0 & 1
  mod <- summary(lm_mod)
  coef = as.data.frame(mod[["coefficients"]])
  coef <- coef %>%
    mutate(lwr=round(Estimate-1.96*`Std. Error`,3),
           upr=round(Estimate+1.96*`Std. Error`,3),
           `95% CI` = paste0("(",lwr,", ",upr,")")) %>%
    rownames_to_column(var = "Term") %>%
    select(-lwr, -upr)%>%
    filter(Term != "(Intercept)")
  print(coef)
}

##Estimate Faith's PD Richness that also takes into account phylogenetic relationship 
library(btools)
library(tidyverse)
#1. Duod
load("Inputs/RData/AMAZE_ps_filtered.RData")
amaze_dup_base <- subset_samples(ps_filtered, Timepoint == "Pre-Intervention")
amaze_dup_base_duod <- subset_samples(amaze_dup_base, Sample.Type == "Duodenal Aspirate")
pd <- estimate_pd(amaze_dup_base_duod)
##combine with metadata 
ad = sample_data(amaze_dup_base_duod)
AlphaDiv_asp = merge(pd, ad, by = 'row.names')

#Implement function
lm_age_duod <- lm_uni_func(equation = PD ~ baseline_age, data = AlphaDiv_asp)
lm_bmi_duod <-lm_uni_func(equation = PD ~ bmi_nustart, data = AlphaDiv_asp)
lm_sex_duod <- lm_uni_func(equation = PD ~ sex_c, data = AlphaDiv_asp)
lm_hiv_duod <- lm_uni_func(equation = PD ~ hiv_status, data = AlphaDiv_asp)
lm_vw_duod <- lm_uni_func(equation = PD ~ VW_mean, data = AlphaDiv_asp)
lm_cd_duod <- lm_uni_func(equation = PD ~ CD_mean, data = AlphaDiv_asp) 
lm_vh_duod <- lm_uni_func(equation = PD ~ VH_mean, data = AlphaDiv_asp)
lm_cle_duod <- lm_uni_func(equation = PD ~ CLE_Fields, data = AlphaDiv_asp)
lm_crp_duod <- lm_uni_func(equation = PD ~ CRP, data = AlphaDiv_asp)
lm_scd14_duod <- lm_uni_func(equation = PD ~ SCD14, data = AlphaDiv_asp)
lm_glp2_duod <- lm_uni_func(equation = PD ~ GLP2, data = AlphaDiv_asp)
lm_lps_duod <- lm_uni_func(equation = PD ~ LPS, data = AlphaDiv_asp)

duod_alpha_stats <- Reduce(function(x,y) merge(x, y, all.x = TRUE, all.y = TRUE),
       list(lm_age_duod, lm_bmi_duod, lm_sex_duod, lm_hiv_duod, lm_vw_duod, lm_cd_duod, lm_vh_duod, lm_cle_duod, lm_crp_duod, lm_scd14_duod, lm_glp2_duod, lm_lps_duod))
write.csv(duod_alpha_stats, "Outputs/2. Baseline/duod_alpha_stats.csv")
rm(list = ls()[grep("^lm", ls())])


#2. Stool
lm_uni_func = function(equation,data){
  lm_mod = lm(formula = equation, data = data) #family must be binomial if y-values are between 0 & 1
  mod <- summary(lm_mod)
  coef = as.data.frame(mod[["coefficients"]])
  coef <- coef %>%
    mutate(lwr=round(Estimate-1.96*`Std. Error`,3),
           upr=round(Estimate+1.96*`Std. Error`,3),
           `95% CI` = paste0("(",lwr,", ",upr,")")) %>%
    rownames_to_column(var = "Term") %>%
    select(-lwr, -upr)%>%
    filter(Term != "(Intercept)")
  print(coef)
}
amaze_dup_base_stool <- subset_samples(amaze_dup_base, Sample.Type == "Stool")
pd <- estimate_pd(amaze_dup_base_stool)
##combine with metadata 
ad = sample_data(amaze_dup_base_stool)
AlphaDiv_stool = merge(pd, ad, by = 'row.names')

#Implement 
lm_age_duod <- lm_uni_func(equation = PD ~ baseline_age, data = AlphaDiv_stool)
lm_bmi_duod <-lm_uni_func(equation = PD ~ bmi_nustart, data = AlphaDiv_stool)
lm_sex_duod <- lm_uni_func(equation = PD ~ sex_c, data = AlphaDiv_stool)
lm_hiv_duod <- lm_uni_func(equation = PD ~ hiv_status, data = AlphaDiv_stool)
lm_vw_duod <- lm_uni_func(equation = PD ~ VW_mean, data = AlphaDiv_stool)
lm_cd_duod <- lm_uni_func(equation = PD ~ CD_mean, data = AlphaDiv_stool) 
lm_vh_duod <- lm_uni_func(equation = PD ~ VH_mean, data = AlphaDiv_stool)
lm_cle_duod <- lm_uni_func(equation = PD ~ CLE_Fields, data = AlphaDiv_stool)
lm_crp_duod <- lm_uni_func(equation = PD ~ CRP, data = AlphaDiv_stool)
lm_scd14_duod <- lm_uni_func(equation = PD ~ SCD14, data = AlphaDiv_stool)
lm_glp2_duod <- lm_uni_func(equation = PD ~ GLP2, data = AlphaDiv_stool)
lm_lps_duod <- lm_uni_func(equation = PD ~ LPS, data = AlphaDiv_stool)

stool_alpha_stats <- Reduce(function(x,y) merge(x, y, all.x = TRUE, all.y = TRUE),
       list(lm_age_duod, lm_bmi_duod, lm_sex_duod, lm_hiv_duod, lm_vw_duod, lm_cd_duod, lm_vh_duod, lm_cle_duod, lm_crp_duod, lm_scd14_duod, lm_glp2_duod, lm_lps_duod))
write.csv(stool_alpha_stats, "Outputs/2. Baseline/stool_alpha_stats.csv")
rm(list = ls()[grep("^lm", ls())])
rm(stool_alpha_stats)
rm(duod_alpha_stats)
rm(ad)
rm(pd)
rm(AlphaDiv_asp)
rm(AlphaDiv_stool)
```

4. Beta Diversity - Bray Curtis distances association with sample characteristics
```{r}
load("Inputs/RData/AMAZE_ps_filtered.RData")
amaze_dup_base <- subset_samples(ps_filtered, Timepoint == "Pre-Intervention")
amaze_dup_base_duod <- subset_samples(amaze_dup_base, Sample.Type == "Duodenal Aspirate")

permanova_func = function(equation,data){
  dist = phyloseq::distance(data, method="bray")
  metadata <- data.frame(sample_data(data))
  require(vegan)
  results <- adonis2(equation, data = metadata, na.action = na.omit)
  coef <- results %>%
    rownames_to_column(var = "Term") %>%
    filter(Term != "Residual") %>%
    filter(Term != "Total")
  print(coef)
}

#1. Duodenal Aspirates
age_duod <- permanova_func(equation = dist ~ baseline_age, data = amaze_dup_base_duod)
bmi_duod <-permanova_func(equation = dist ~ bmi_nustart, data = amaze_dup_base_duod)
sex_duod <- permanova_func(equation = dist ~ sex_c, data = amaze_dup_base_duod)
hiv_duod <- permanova_func(equation = dist ~ hiv_status, data = amaze_dup_base_duod)
vw_duod <- permanova_func(equation = dist ~ VW_mean, data = amaze_dup_base_duod)
cd_duod <- permanova_func(equation = dist ~ CD_mean, data = amaze_dup_base_duod) 
vh_duod <- permanova_func(equation = dist ~ VH_mean, data = amaze_dup_base_duod)
cle_duod <- permanova_func(equation = dist ~ CLE_Fields, data = amaze_dup_base_duod)
glp2_duod <- permanova_func(equation = dist ~ GLP2, data = amaze_dup_base_duod)
scd14_duod <- permanova_func(equation = dist ~ SCD14, data = amaze_dup_base_duod)
crp_duod <- permanova_func(equation = dist ~ CRP, data = amaze_dup_base_duod)
lps_duod <- permanova_func(equation = dist ~ LPS, data = amaze_dup_base_duod)

duod_beta_stats <- Reduce(function(x,y) merge(x, y, all.x = TRUE, all.y = TRUE),
       list(age_duod, bmi_duod, sex_duod, hiv_duod, vw_duod, cd_duod, vh_duod, cle_duod, glp2_duod,
            scd14_duod, crp_duod, lps_duod))
write.csv(duod_beta_stats, "Outputs/2. Baseline/duod_beta_stats.csv")

#2. Stool
amaze_dup_base_stool <- subset_samples(amaze_dup_base, Sample.Type == "Stool")

age_stool <- permanova_func(equation = dist ~ baseline_age, data = amaze_dup_base_stool)
bmi_stool <-permanova_func(equation = dist ~ bmi_nustart, data = amaze_dup_base_stool)
sex_stool <- permanova_func(equation = dist ~ sex_c, data = amaze_dup_base_stool)
hiv_stool <- permanova_func(equation = dist ~ hiv_status, data = amaze_dup_base_stool)
vw_stool <- permanova_func(equation = dist ~ VW_mean, data = amaze_dup_base_stool)
cd_stool <- permanova_func(equation = dist ~ CD_mean, data = amaze_dup_base_stool) 
vh_stool <- permanova_func(equation = dist ~ VH_mean, data = amaze_dup_base_stool)
cle_stool <- permanova_func(equation = dist ~ CLE_Fields, data = amaze_dup_base_stool)
glp2_stool <- permanova_func(equation = dist ~ GLP2, data = amaze_dup_base_stool)
scd14_stool <- permanova_func(equation = dist ~ SCD14, data = amaze_dup_base_stool)
crp_stool <- permanova_func(equation = dist ~ CRP, data = amaze_dup_base_stool)
lps_stool <- permanova_func(equation = dist ~ LPS, data = amaze_dup_base_stool)

stool_beta_stats <- Reduce(function(x,y) merge(x, y, all.x = TRUE, all.y = TRUE),
       list(age_stool, bmi_stool, sex_stool, hiv_stool, vw_stool, cd_stool, vh_stool, cle_stool,
            glp2_stool, scd14_stool, crp_stool, lps_stool))
write.csv(stool_beta_stats, "Outputs/2. Baseline/stool_beta_stats.csv")
```

5. Baseline microbiome models with baseline biomarkers and morphometry
```{r}
load("Inputs/RData/AMAZE_ps_filtered.RData")
lm_mod_func = function(var,data){
  require(microViz)
  mod_lm <- data %>%
  tax_fix() %>%
  tax_transform("compositional", rank = "Genus") %>%
  tax_filter(min_prevalence = 0.1, undetected = 0, use_counts = T) %>%
  taxatree_models(type = "lm", 
                  rank = "Genus",
                  trans = "log2", 
                  trans_args = list(zero_replace = "halfmin"),
                  variables = c(var, "hiv_status", "baseline_age"))
  lm_stats <- taxatree_models2stats(mod_lm)
  lm_stats <- taxatree_stats_p_adjust(data = lm_stats, method = "BY", grouping = "rank")
  lm_stats2 <- lm_stats %>% taxatree_stats_get() %>% filter(p.adj.BY.rank < 0.05) %>% filter(term == var)
  print(lm_stats2)
}

#1. Duodenal samples
amaze_dup_base <- subset_samples(ps_filtered, Timepoint == "Pre-Intervention")
amaze_dup_base_duod <- subset_samples(amaze_dup_base, Sample.Type == "Duodenal Aspirate")
crp_lm <- lm_mod_func(var = "CRP", data = amaze_dup_base_duod)
scd14_lm <- lm_mod_func(var = "SCD14", data = amaze_dup_base_duod)
lps_lm <- lm_mod_func(var = "LPS", data = amaze_dup_base_duod)
glp2_lm <- lm_mod_func(var = "GLP2", data = amaze_dup_base_duod)
vh_lm <- lm_mod_func(var = "VH_mean", data = amaze_dup_base_duod)
cd_lm <- lm_mod_func(var = "CD_mean", data = amaze_dup_base_duod)
vw_lm <- lm_mod_func(var = "VW_mean", data = amaze_dup_base_duod)
cle_lm <- lm_mod_func(var = "CLE_Fields", data = amaze_dup_base_duod)
#merge the data and save
df_list <- list(crp_lm, scd14_lm, lps_lm, vh_lm, cd_lm, vw_lm, cle_lm)
lm_biomarkers_duod <- reduce(df_list,full_join)
write.csv(lm_biomarkers_duod, "Outputs/2. Baseline/biomarker_morph_assoc_duod.csv")


#2. Stool
amaze_dup_base_stool <- subset_samples(amaze_dup_base, Sample.Type == "Stool")
crp_lm <- lm_mod_func(var = "CRP", data = amaze_dup_base_stool)
scd14_lm <- lm_mod_func(var = "SCD14", data = amaze_dup_base_stool)
lps_lm <- lm_mod_func(var = "LPS", data = amaze_dup_base_stool)
glp2_lm <- lm_mod_func(var = "GLP2", data = amaze_dup_base_stool)
vh_lm <- lm_mod_func(var = "VH_mean", data = amaze_dup_base_stool)
cd_lm <- lm_mod_func(var = "CD_mean", data = amaze_dup_base_stool)
vw_lm <- lm_mod_func(var = "VW_mean", data = amaze_dup_base_stool)
cle_lm <- lm_mod_func(var = "CLE_Fields", data = amaze_dup_base_stool)
#merge the data and save
df_list <- list(crp_lm, scd14_lm, lps_lm, vh_lm, cd_lm, vw_lm, cle_lm)
lm_biomarkers_stool <- reduce(df_list,full_join)
write.csv(lm_biomarkers_stool, "Outputs/2. Baseline/biomarker_morph_assoc_stool.csv")
```

6. Differential abundance of genera  for HIV status in stool samples - Maaslin2
```{r}
#get data
load("Inputs/RData/AMAZE_ps_filtered.RData")
ps_stool <- subset_samples(ps_filtered, Sample.Type == "Stool")
ps_stool.pre <- subset_samples(ps_stool, Timepoint == "Pre-Intervention")
library(Maaslin2)
mas_P <- Maaslin2(
  input_data = data.frame(otu_table(ps_stool.pre)),
  input_metadata = data.frame(sample_data(ps_stool.pre)),
  output = "Outputs/2. Baseline/DiffStool_HIV_Status",
  min_abundance = 0.001,
  min_prevalence = 0.05,
  normalization = "TMM",
  transform = "NONE",
  analysis_method = "NEGBIN",
  max_significance = 0.1,
  fixed_effects = c("hiv_status"),
  reference = c("hiv_status,Negative"),
  correction = "BY",
  standardize = FALSE,
  cores = 1,
  plot_heatmap = TRUE,
  plot_scatter = TRUE,
  heatmap_first_n = 50)
pp.stool <- mas_P[["results"]]
SigBact <- pp.stool %>%
  mutate(sam.type = ifelse(.$coef > 0,  "Positive", "Negative"),
         coef.exp = exp(coef),
         logfold2 = log2(coef.exp),
         logfold2.2 = ifelse(logfold2 < 0, logfold2*-1, logfold2)) %>%
  filter(qval < 0.05)

lda.hivstatus <- ggpubr::ggbarplot(SigBact, x = "feature", y = "logfold2.2",
          fill = "sam.type",           # change fill color by mpg_level
          color = "white",            # Set bar border colors to white
          palette = c("blue", "red"),            # jco journal color palett. see ?ggpar
          sort.val = "asc",           # Sort the value in ascending order
          sort.by.groups = T,     # Don't sort inside each group
          x.text.angle = 90,          # Rotate vertically x axis texts
          ylab = "Log2Fold Change",
          xlab = "Taxon",
          legend.title = "HIV Status",
          rotate = TRUE,
          ggtheme = theme_classic()
          ) 
ggsave("Outputs/2. Baseline/SuppFig4.png", plot = lda.hivstatus, dpi = 600, device = "png", width = 18, height = 15, units = "cm")
```

Duodenal and Stool microbial correlations
```{r}

load("Inputs/RData/AMAZE_ps_filtered.RData")

library(ggplot2)
library(phyloseq)
library(tidyverse)
library(microbiome)
library(microViz)
amaze_dup2 <- subset_samples(ps_filtered, Timepoint == "Pre-Intervention")

tax_t <- as.data.frame(amaze_dup2@otu_table) 
met_t <- as.data.frame(amaze_dup2@sam_data)


tax_2 <- tax_t %>%
  rownames_to_column(var = "Sample")
q <- ps_melt(amaze_dup2)
qs <- q%>%
  select(Sample, PID, InterventionArm, Timepoint, hiv_status, sex_c, Sample.Type) %>%
  #filter( %>%
  unique()

for_plot <- left_join(qs, tax_2)
for_plot2 <- for_plot %>%
  select(-Sample) %>%
  pivot_wider(names_from = Sample.Type, 
              values_from = -c(PID, InterventionArm, Timepoint, hiv_status, sex_c, Sample.Type),
              names_glue = "{.value}_{Sample.Type}")
for_plot2[is.na(for_plot2)] <- 0

library(corrplot)
for_corr <- for_plot2 %>%
  select(-c(PID, InterventionArm, Timepoint, hiv_status, sex_c)) %>%
  select(where(~ sum(.) != 0))


#rename taxa
num_rows <- 394
seq_let <- rep (letters, length.out = num_rows)

forcorrr_t <- t(for_corr)
forcorr2 <- forcorrr_t %>%
  as.data.frame()%>%
  rownames_to_column(var= "tax") %>%
  mutate(let_col = paste0("tax_",1:n()))
write.csv(forcorr2 %>%
            select(tax, let_col), "Outputs/2. Baseline/SpearmanCorr/microbe_names.csv")

forcorr3 <- forcorr2 %>%
  select(-tax)  %>%
  column_to_rownames(var = "let_col")
forcorr2_t <- t(forcorr3)  

#Calculate correlation coefficient
library(psych)
library(tidyverse)
adj.dat <- corr.test(forcorr2_t, method = "spearman", adjust = "holm")
write.csv(adj.dat$ci, "Outputs/2. Baseline/SpearmanCorr/microbe_cor_pval.csv")
write.csv(adj.dat$ci2, "Outputs/2. Baseline/SpearmanCorr/microbe_cor_pval_adjust.csv")
write.csv(adj.dat$r, "Outputs/2. Baseline/SpearmanCorr/microbe_cor_rvalues.csv")

dat3 <- adj.dat$ci %>%
  rownames_to_column(var = "tax0") %>%
  separate(tax0, into = c("tax1", "tax2"), sep = "-") %>%
  separate(tax1, into = c("tx01", "tx1"), sep = "_") %>%
  separate(tax2, into = c("tx02", "tx2"), sep = "_") %>%
  select(-tx01, -tx02) %>%
  filter(p < 0.05) %>%
  filter(p != 0)

dd2 <- forcorr2 %>%
  select(tax, let_col) %>%
  separate(let_col, into = c("tx01", "tx1"), sep = "_") %>%
  select(-tx01)

data_fin <- merge(dat3, dd2, by = "tx1", all = T)

dd3 <- forcorr2 %>%
  select(tax, let_col) %>%
  separate(let_col, into = c("tx01", "tx2"), sep = "_") %>%
  select(-tx01) %>%
  rename(tax2 = tax)
data_fin2 <- merge(data_fin, dd3, by = "tx2", all = T)

term1 <- "Stool"
term2 <- "Duodenal Aspirate"
data_fin2 <- data_fin2%>%
  filter(!is.na(tx2)) %>%
  filter((str_detect(tax, term1) & str_detect(tax2, term2)) |
           (str_detect(tax, term2) & str_detect(tax2, term1)))

write.csv(data_fin2, "Outputs/2. Baseline/microbe_cor_sig_duod_stool.csv")
```

